---
title: "Deletions - Repeatability"
author: "Joe Shaw (CS20980)"
date: today
date-format: "DD/MM/YYYY"
format: pdf
editor: visual
---

# Aim

The aim of this experiment is to determine the uncertainty of measurement for
deletion calling on PanSolid.

```{r}
#| label: packages-and-functions
#| include: FALSE

library(tidyverse)
library(readxl)
library(here)
library(janitor)

source(here("functions/pansolid_excel_functions.R"))
source(here("functions/uom_functions.R"))

```

# Data

I have used the initial outputs from the CLC pipeline.

```{r}
#| label: data
#| include: FALSE

bio_cnv_folder <- "S:/central shared/Genetics/NGS/Bioinformatics/1_Pan-solid-Cancer/CNV/Deletions/"

data_folder <- config::get("data_filepath")

del_files <- list.files(path = bio_cnv_folder,
                        recursive = FALSE,
                        full.names = TRUE,
                        pattern  = "TSG\\s\\(Deleted\\).*.xlsx")

del_genes <- read_excel(path = paste0(data_folder,
                                    "validation/DOC6567_deletions/gene_lists/",
                                    "2024_12_20_deletion_gene_list.xlsx")) |> 
  janitor::clean_names() |> 
  select(priority, gene)

read_all_targets <- function(filepath, sheet_no) {
  
  df <- read_excel(path = filepath,
                   sheet = sheet_no,
                   col_types = c("text", "text", "text", "numeric",
                                 "numeric", "numeric", "numeric", 
                                 "numeric", "numeric", "numeric", "numeric", 
                                 "numeric", "numeric", "text", "text", 
                                 "numeric", "numeric", "text", 
                                 "text", "text")) |> 
    mutate(filepath = filepath) |> 
    janitor::clean_names()
  
  output <- add_identifiers(filepath, df)
  
  return(output)
  
}

coarse_target_df <- del_files |> 
  map(\(del_files) read_all_targets(del_files,
                                    sheet_no = 3)) |> 
  list_rbind() |> 
  mutate(graining = "coarse")

fine_target_df <- del_files |> 
  map(\(del_files) read_all_targets(del_files,
                                    sheet_no = 4)) |> 
  list_rbind() |> 
  mutate(graining = "fine")

all_target_df <- rbind(coarse_target_df, fine_target_df)

target_df <- read_csv(paste0(data_folder,
                             "validation/DOC6283_amplifications/bed_files/",
                             "PanSolidv2_GRCh38_noalt_BED.csv")) |> 
  clean_names() |> 
  mutate(target_type = case_when(
    
    grepl(x = name, 
          pattern =  "chr(\\d{1,2}|X):\\d{1,3}.+") == TRUE ~"genomic backbone",
    TRUE ~"gene target"
  ))

target_df_with_coordinates <- extract_pansolid_cnv_coordinates(target_df, 
                                                               region) |> 
  mutate(target_size_bp = abs(end-start))

gene_target_counts <- target_df |> 
  filter(name %in% del_genes$gene | name == "DICER1") |> 
  count(name) |>
  mutate("CLC setting" = case_when(
    n < 10 ~"Fine",
    n >= 10 ~"Coarse"
  )) |> 
  janitor::clean_names()

genes_on_coarse <- gene_target_counts |> 
  filter(clc_setting == "Coarse")

genes_on_fine <- gene_target_counts |> 
  filter(clc_setting == "Fine")

```

# Samples

I have focussed on the 3 samples (24017319, 24027774 and 24020998) which were 
repeated on multiple worksheets as part of the PanSolid amplifications 
validation. Of these, 24017319 has results from whole genome sequencing.

```{r}
#| label: replicates
#| include: FALSE

inter_run_replicates <- c(
  # 24017319
  "24017319_WS141734", "24017319a_WS144291", "24017319_WS144364",
  # 24027774
  "24027774_WS142076", "24027774a_WS144291", "24027774_WS144364",
  # 24020998
  "24020998_WS141045", "24020998a_WS144291", "24020998_WS144364")

intra_run_replicates <- c(
  # 24017319
  "24017319a_WS144291", "24017319b_WS144291", 
  # This replicate removed because of low quality
  # "24017319c_WS144291",
  
  # 24027774
  "24027774a_WS144291", "24027774b_WS144291", "24027774c_WS144291",
  # 24020998
  "24020998a_WS144291", "24020998b_WS144291", "24020998c_WS144291"
)

```

# Genes

I have filtered the target fold changes for each sample to focus on the `r length(del_genes$gene)` genes with priority 1 or priority 2.

```{r}
#| label: del-genes
#| echo: FALSE

print(del_genes$gene)

```

# Uncertainty of Measurement

For the uncertainty of measurement analysis I've filtered the PanSolid results
by graining level: genes with 10 or more targets are analysed on "coarse" and 
genes with fewer than 10 targets are analysed on "fine".

I have also only used dosage results for the `r length(del_genes$gene)` genes on
the deletions list.

```{r}
#| label: identify-repeat-results
#| include: FALSE

repeat_results <- all_target_df |> 
  filter((name %in% genes_on_coarse$name & graining == "coarse") |
           (name %in% genes_on_fine$name & graining == "fine")) |> 
  filter(labno_suffix_worksheet %in% inter_run_replicates |
           labno_suffix_worksheet %in% intra_run_replicates) |> 
  filter(name %in% del_genes$gene) 

```

I then performed the following steps:

- Converted any fold changes calculated with the "gain" fold change equation
 to the results for the "loss" equation.
 
- Took the lowest regional fold change for each gene (some genes had different
regions at different fold changes)

- Removed results for the CDKN2A, CDKN2B and PTEN genes where the fold change
was below -1.75. This was because these calls were for homozygous deletions
and subsequently have more variation than regions with normal or single loss
dosage.

```{r}
#| label: select-uom-data
#| include: FALSE

del_fc_threshold <- -1.33333

repeat_data_filtered <- repeat_results |> 
  distinct(worksheet, labno, suffix,
           labno_suffix_worksheet, 
           filepath, name, region_joined_targets, regional_fold_change, 
           regional_consequence, graining) |> 
  mutate(worksheet_suffix = str_c(worksheet, "_", suffix)) |> 
  # Convert amplification fold changes to deletion fold changes
  mutate(fold_change_uom = case_when(
    regional_fold_change >= 1 ~-1 / regional_fold_change,
    TRUE ~regional_fold_change)) |> 
  group_by(labno_suffix_worksheet, name) |> 
  summarise(min_fold_change = min(fold_change_uom)) |> 
  mutate(pansolid_call = case_when(
      min_fold_change <= -1.3333 ~"del",
      min_fold_change > -1.3333 ~"normal"),
      labno = str_extract(labno_suffix_worksheet,
                             pattern = "\\d{8}")) 

```

Here is the filtered data without the CDKN2A and PTEN homozygous deletions
removed.

```{r}
#| label: plot-uom-data
#| echo: FALSE

ggplot(repeat_data_filtered, aes(x = name, 
                                    y = min_fold_change)) +
  geom_point(shape = 21) +
  theme_bw() +
  theme(axis.text.x = element_text(angle =90)) +
  facet_wrap(~labno, nrow = 3) +
  geom_hline(yintercept = -1.33, linetype = "dashed") 


```

```{r}
#| label: calculate-uom
#| include: FALSE

repeat_data_uom <- repeat_data_filtered  |> 
  filter(!(name %in% c("CDKN2A", "CDKN2B",
                       "PTEN") & min_fold_change < -1.75))

inter_sd_table <- group_sd(repeat_data_uom |> 
                             filter(labno_suffix_worksheet %in% inter_run_replicates),
                           labno, name, min_fold_change)

inter_sd <- round(pool_sd(inter_sd_table), 3)

intra_sd_table <- group_sd(repeat_data_uom |> 
                             filter(labno_suffix_worksheet %in% intra_run_replicates),
                           labno, name, min_fold_change)

intra_sd <- round(pool_sd(intra_sd_table), 3)

all_sd_table <- group_sd(repeat_data_uom,
                         labno, name, min_fold_change)

all_sd <- round(pool_sd(all_sd_table), 3)

del_uom <- define_uom(max(inter_sd, intra_sd, all_sd), 2)

```

{{< pagebreak >}}

I calculated the pooled standard deviation for the fold change of each gene in 
the 3 samples.

**Intra-run standard deviation: `r intra_sd`**

**Inter-run standard deviation: `r inter_sd`**

**Total standard deviation: `r all_sd`**

For comparison, the intra, inter and total standard deviation for amplifications
were 0.032, 0.085 and 0.077.

Taking the highest standard deviation value, the uncertainty of measurement 
can be calculated.

**Uncertainty of measurement for fold change: ±`r del_uom` (95% confidence)**

This would mean a deletion fold change threshold of 
`r del_fc_threshold + del_uom`.

This is what adding the upper threshold onto the repeat data would look like.

```{r}
#| label: new-threshold
#| echo: FALSE

ggplot(repeat_data_filtered, aes(x = name, 
                                    y = min_fold_change)) +
  geom_point(shape = 21) +
  theme_bw() +
  theme(axis.text.x = element_text(angle =90)) +
  facet_wrap(~labno, nrow = 3) +
  geom_hline(yintercept = del_fc_threshold, linetype = "dashed") +
  geom_hline(yintercept = del_fc_threshold + del_uom, linetype = "dashed") +
  coord_cartesian(ylim = c(-1.75, -0.25))

```
