---
title: "Deletions - Repeatability"
author: "Joe Shaw (CS20980)"
format: pdf
editor: visual
---

# Aim

The aim of this experiment is to determine the uncertainty of measurement for
deletion calling on PanSolid.

```{r}
#| label: packages-and-functions
#| include: FALSE

library(tidyverse)
library(readxl)
library(here)

source(here("functions/pansolid_excel_functions.R"))
source(here("functions/uom_functions.R"))

```

# Data

I have used the initial outputs from the CLC pipeline.

```{r}
#| label: data
#| include: FALSE

bio_cnv_folder <- "S:/central shared/Genetics/NGS/Bioinformatics/1_Pan-solid-Cancer/CNV/Deletions/"

data_folder <- config::get("data_filepath")

del_files <- list.files(path = bio_cnv_folder,
                        recursive = FALSE,
                        full.names = TRUE,
                        pattern  = "TSG\\s\\(Deleted\\).*.xlsx")

del_genes <- read_csv(file = paste0(data_folder,
                                    "validation/DOC6567_deletions/gene_lists/",
                                    "del_genes_new.csv"))

read_all_targets <- function(filepath, sheet_no) {
  
  df <- read_excel(path = filepath,
                   sheet = sheet_no,
                   col_types = c("text", "text", "text", "numeric",
                                 "numeric", "numeric", "numeric", 
                                 "numeric", "numeric", "numeric", "numeric", 
                                 "numeric", "numeric", "text", "text", 
                                 "numeric", "numeric", "text", 
                                 "text", "text")) |> 
    mutate(filepath = filepath) |> 
    janitor::clean_names()
  
  output <- add_identifiers(filepath, df)
  
  return(output)
  
}

coarse_target_df <- del_files |> 
  map(\(del_files) read_all_targets(del_files,
                                    sheet_no = 3)) |> 
  list_rbind() |> 
  mutate(graining = "coarse")

fine_target_df <- del_files |> 
  map(\(del_files) read_all_targets(del_files,
                                    sheet_no = 4)) |> 
  list_rbind() |> 
  mutate(graining = "fine")

all_target_df <- rbind(coarse_target_df, fine_target_df)

```

# Samples

I have focussed on the 3 samples (24017319, 24027774 and 24020998) which were
repeated on multiple worksheets as part of the PanSolid amplifications validation.
Of these, 24017319 has results from whole genome sequencing.

```{r}
#| label: replicates
#| include: FALSE

inter_run_replicates <- c(
  # 24017319
  "24017319_WS141734", "24017319a_WS144291", "24017319_WS144364",
  # 24027774
  "24027774_WS142076", "24027774a_WS144291", "24027774_WS144364",
  # 24020998
  "24020998_WS141045", "24020998a_WS144291", "24020998_WS144364")

intra_run_replicates <- c(
  # 24017319
  "24017319a_WS144291", "24017319b_WS144291", 
  # This replicate removed because of low quality
  # "24017319c_WS144291",
  
  # 24027774
  "24027774a_WS144291", "24027774b_WS144291", "24027774c_WS144291",
  # 24020998
  "24020998a_WS144291", "24020998b_WS144291", "24020998c_WS144291"
)

```

# Genes

I have filtered the target fold changes for each sample to focus on the `r length(del_genes$gene)` genes specified by George and Helene. The full gene list is:

```{r}
#| label: del-genes
#| echo: FALSE

print(del_genes$gene)

```

# Uncertainty of Measurement

```{r}
#| label: calculate-uom
#| include: FALSE

repeat_results <- coarse_target_df |> 
  filter(labno_suffix_worksheet %in% inter_run_replicates |
           labno_suffix_worksheet %in% intra_run_replicates) |> 
  filter(name %in% del_genes$gene) 

del_fc_threshold <- -1.33333

repeat_results_distinct <- repeat_results |> 
  distinct(worksheet, labno, suffix,
           labno_suffix_worksheet, 
           filepath, name, region_joined_targets, regional_fold_change, 
           regional_consequence, graining) |> 
  mutate(worksheet_suffix = str_c(worksheet, "_", suffix),
         pansolid_call_minus_1_3 = case_when(
           regional_fold_change <= del_fc_threshold ~"Deletion",
           regional_fold_change > del_fc_threshold ~"Normal"
         ))

repeat_data_uom <- repeat_results_distinct |> 
  # Convert amplification fold changes to deletion fold changes
  mutate(fold_change_uom = case_when(
    regional_fold_change >= 1 ~-1 / regional_fold_change,
    TRUE ~regional_fold_change))

inter_sd_table <- group_sd(repeat_data_uom |> 
                             filter(labno_suffix_worksheet %in% inter_run_replicates),
                           labno, name, fold_change_uom)

inter_sd <- round(pool_sd(inter_sd_table), 3)

intra_sd_table <- group_sd(repeat_data_uom |> 
                             filter(labno_suffix_worksheet %in% intra_run_replicates),
                           labno, name, fold_change_uom)

intra_sd <- round(pool_sd(intra_sd_table), 3)

all_sd_table <- group_sd(repeat_data_uom,
                         labno, name, fold_change_uom)

all_sd <- round(pool_sd(all_sd_table), 3)

del_uom <- define_uom(max(inter_sd, intra_sd, all_sd), 2)

```

I calculated the pooled standard deviation for the fold change of each gene in 
the 3 samples.

**Intra-run standard deviation: `r intra_sd`**

**Inter-run standard deviation: `r inter_sd`**

**Total standard deviation: `r all_sd`**

The standard deviation is greater for deletions than amplifications. The intra, 
inter and total standard deviation for amplifications were 0.032, 0.085 and 0.077.

Taking the highest standard deviation value, the uncertainty of measurement can be calculated.

**Uncertainty of measurement for fold change: ±`r del_uom` (95% confidence)**

This would mean a deletion fold change threshold of `r del_fc_threshold + del_uom`.

```{r}
#| label: new-threshold
#| include: FALSE

repeat_data_uom_mod <- repeat_data_uom |> 
  mutate(pansolid_call_new_threshold = case_when(
    regional_fold_change <=  del_fc_threshold + del_uom ~"Deletion",
    regional_fold_change >  del_fc_threshold + del_uom ~"Normal"
  ))


make_repeat_plot <- function(labno, graining_level, 
                             threshold) {
  
  plot <- repeat_data_uom_mod |> 
    filter(labno == {{ labno }} & graining == {{ graining_level }}) |> 
    ggplot(aes(x = worksheet_suffix,
               y = regional_fold_change)) +
    annotate('rect', xmin = 0, xmax = 5, ymin = -1, 
             ymax = 1, fill='white') +
    geom_hline(yintercept = del_fc_threshold, linetype = "dashed") +
    geom_hline(yintercept = 1, colour = "grey") +
    geom_hline(yintercept = -1, colour = "grey") +
    geom_point(shape = 21, size = 2, aes(fill = {{ threshold }})) +
    scale_fill_manual(values = c("red", "white")) +
    theme_bw() +
    theme(axis.text.x = element_blank()) +
    facet_wrap(~name) +
    labs(x = "Replicate", y = "Fold change", fill = "",
         title = labno)
  
  return(plot)
  
}

```

{{< pagebreak >}}

```{r}
#| label: fig-24017319-original-threshold
#| fig-cap: "24017319 original results"
#| fig-width: 6
#| fig-height: 7
#| echo: FALSE

make_repeat_plot("24017319", "coarse", pansolid_call_minus_1_3) +
  coord_cartesian(ylim = c(-3, 3)) +
  labs(title = "24017319 -1.33 threshold")

```

{{< pagebreak >}}

```{r}
#| label: fig-24017319-new-threshold
#| fig-cap: "24017319 results with new threshold"
#| fig-width: 6
#| fig-height: 7
#| echo: FALSE

make_repeat_plot("24017319", "coarse", pansolid_call_new_threshold) +
  coord_cartesian(ylim = c(-3, 3)) +
  labs(title = "24017319 -1.06 threshold")

```
