---
title: "DOC6791: Validation of Detecting Chromosome Level Copy Number Variants with the PanSolid CLC pipeline"
format: 
  docx:
    reference-doc: north_west_glh_document_template.docx
    link-citations: true
toc: true
toc-title: "Table of Contents"
bibliography: pansolid_cnv_references.bib
csl: harvard-manchester-metropolitan-university.csl
fig-align: "left"
fig-width: 6
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: packages-and-functions
#| include: FALSE

library(tidyverse)
library(janitor)
library(here)
library(readxl)

source(here("functions/extract_pansolid_cnv_coordinates.R"))
source(here("functions/pansolid_cnv_excel_functions.R"))
source(here("functions/chromosome_arm_functions.R"))

```

# Authors

Author: Joe Shaw (CS20980)

Project Lead(s): Joe Shaw, Elizabeth Sollars, Eleanor Baker, Helene
Schlecht, George Burghel

Date: `r Sys.Date()`

Author contact: joseph.shaw2\@mft.nhs.uk; joseph.shaw3\@nhs.net

# Acronyms

**ATRX**: α-thalassemia/mental retardation syndrome X-linked

**cIMPACT-NOW:** consortium to Inform Molecular and Practical Approaches
to CNS Tumour taxonomy-Not Official WHO

**CLC**: there is no official acronym for CLC, although unofficially it was referred to as "Cake Loving Company" by CLC employees [@sollars2025]

**DNA**: DeoxyriboNucleic Acid

**FFPE**: Formalin-Fixed Paraffin Embedded

**FISH**: Fluorescent In-Situ Hybridisation

**GLH**: Genomic Laboratory Hub

**IDH**: isocitrate dehydrogenase

**IHC**: Immuno-HistoChemistry

**LOH**: Loss Of Heterozygosity

**MGMT**: O6-Methylguanine-DNA methyltransferase

**NCC**: Neoplastic Cell Content

**NGS**: Next Generation Sequencing

**NHS**: National Health Service

**PCR:** Polymerase Chain Reaction

**SNP:** Single Nucleotide Polymorphism

**TERT**: Telomerase Reverse Transcriptase

**WGS**: Whole Genome Sequencing

**WHO**: World Health Organisation

# Introduction

## Purpose

The purpose of this document is to describe the validation of detection
of chromosome level copy number variants (CNVs) in solid tumour samples
with the PanSolid v2 QIAseq enrichment panel using the PanSolid CLC
pipeline (v2.4) at the North West Genomic Laboratory Hub (GLH).
Detection of specific CNVs in different cancer types is required by the
most recent update of the
[NHS Genomic Test Directory for
Cancer](https://www.england.nhs.uk/publication/national-genomic-test-directories/).

## Target copy number variants

### Neurological tumours

Co-deletion of 1p and 19q, as determined by PCR-based LOH testing, was
shown to be associated with a good response to chemotherapy in
oligodendrogliomas [@cairncross1998]. The biological mechanism for this
observation remains unknown.

Later cytogenetic studies using Giemsa-banding and FISH showed that the
co-deletion of 1p and 19q was due to a recurrent translocation between
the two chromosomes. A derivative der(1;19)(q10;p10) chromosome was
observed within glioma cells, whilst the reciprocal der(1;19)(p10;q10)
chromosome is absent and presumably lost [@jenkins2006; @griffin2006].

In the 2021 edition of the World Health Organisation Guidelines,
detection of 1p19q codeletion is required for classification of a tumour
as "Oligodendroglioma, IDH-mutant, and 1p/19q-codeleted" [@louis2021].
Critically, 1p19q codeletion is a key biomarker for distinguishing
between oligodendrogliomas, which contain the codeletion, and
astrocytomas, which do not [@masui2016].

Multiple techniques can be used to detect 1p19q codeletions and there is
no gold-standard method [@brandner2022].

In the most recent update of the cIMPACT-NOW guidelines for meningiomas,
whole chromosome copy number analysis is recommended for tumour grading,
with at least two losses of chromosomes 6, 10, 14 or 18 sufficient to
categorise a tumour as molecular high-risk [@sahm2025].

### Malignant melanoma

The purpose is to distinguish benign nevi from malignant melanomas.

References: @ferrara2016; @gerami2009

### Renal cell carcinoma

Specific chromosome gains and losses are associated with subtypes of
renal cancer.

Renal cell carcinoma: loss of 3p (including VHL)

Papillary renal cell carcinoma: gains of 7 and 17

References: @liu2020, @klatte2009, @chen2009

## PanSolid enrichment

```{r}
#| label: enrichment-targets
#| include: FALSE

target_df <- read_csv(paste0(config::get("data_folderpath"),
                             "validation/DOC6283_amplifications/bed_files/",
                             "PanSolidv2_GRCh38_noalt_BED.csv")) |> 
  clean_names() |> 
  mutate(chromosome_fct = factor(chromosome,
                                levels  = c(seq(1:22), "X")))

target_df_coordinates <- target_df |> 
  extract_pansolid_cnv_coordinates(cnv_coord_col = region) |> 
  mutate(chromosome_char = as.character(chromosome_fct)) |> 
  add_chromosome_arms()

target_chr_arm_summary <- target_df_coordinates |> 
  filter(!is.na(chromosome_arm)) |> 
  group_by(chromosome_fct, chromosome_arm) |> 
  summarise(min_coord = min(start),
            max_coord = max(end)) |> 
  mutate(target_span = max_coord - min_coord) |> 
  left_join(chr_coords_grch38,
            join_by("chromosome_arm" == "chrom_arm")) |> 
  mutate(target_percent = (target_span / length)*100,
         p_or_q = str_extract(string = chromosome_arm,
                              pattern = "p|q"))
  
targets_per_chromosome <- target_df_coordinates |> 
  filter(!is.na(chromosome_arm)) |> 
  group_by(chromosome_fct, chromosome_arm) |> 
  count() |> 
  mutate(p_or_q = str_extract(string = chromosome_arm,
                              pattern = "p|q"))

```

```{r}
#| label: fig-targets-per-chromosome
#| echo: FALSE

fig_targets_per_chromosome <- ggplot(targets_per_chromosome, 
                                     aes(x = chromosome_fct, y = n)) +
  geom_col(aes(fill = p_or_q),
           colour = "black") +
  theme_bw() +
  labs(x = "Chromosome", y = "Number of targets",
       title = "Distribution of targets on PanSolid enrichment",
       fill = "")

fig_targets_per_chromosome

```

```{r}
#| label: fig-target-chr-percent
#| echo: FALSE

fig_target_chr_percent <- ggplot(target_chr_arm_summary, 
                                 aes(x = chromosome_fct, y = target_percent)) +
  geom_col(position = "dodge", 
           colour = "black",
           aes(fill = p_or_q)) +
  facet_wrap(~p_or_q, ncol = 1) +
  theme_bw() +
  labs(fill = "",
       x = "Chromosome",
       y = "Percentage covered by PanSolid targets")

fig_target_chr_percent

```

## Requirements and Performance Specification

The requirements for this new method are as follows:

-   The test must have a clear visual presentation of results
-   The test must provide a quantitative result for whether a chromosome
    or chromosome arm is impacted by a gain or loss.
-   The combined analysis of the PanSolid CLC pipeline with manual
    review by a clinical scientist should sensitivity and specificity
    values above 90% when compared with orthogonal testing.
-   The test must have clear quality metric thresholds for deciding
    sample quality

# Methods

## Samples

```{r}
#| label: load-cohort
#| include: FALSE

val_cohort <- read_excel(path = paste0(config::get("data_folderpath"),
                                       "validation/DOC6791_chromosome_arms/",
                                       "chromosome_arm_validation_cohort.xlsx"))
```

A cohort of 
`r length(unique(val_cohort$pansolid_labno))`
different samples was assembled for this validation.

## Next generation sequencing

Next generation sequencing on the PanSolid v2 enrichment was performed
for all samples.

## PanSolid CLC bioinformatics pipeline

The PanSolid CNV pipeline was built in CLC Genomics Workbench and has been described in previous validation documents.

### Visualisation and quantification

The outputs of the PanSolid CNV pipeline are visualised at the whole-genome level via interactive plots, which were generated using the plotly package by Elizabeth Sollars and implemented during the validation of deletion and loss of heterozygosity detection (DOCxxxx).
In addition, scientists in the somatic cancer team requested an quantitative way of alerting them when a CNV impacting a whole chromosome arm or arms was present, such as the 1p19q co-deletion.
The following method 

1.  Load the ploidy regions from the "Ploidy Regions" tab of the "Results_SNVs_Indels" Excel file in the "Unannotated_Files" folder of each worksheet.

2.  Annotate each region by the chromosome arm that the region is on.

3.  Calculate the size of the region by subtracting the start coordinate from the end coordinate.

4.  Sum the total length, in base pairs, of ploidy regions for a particular ploidy state, such as "Deletion", on each chromosome arm.

5.  Calculate the percentage of the entire chromosome arm that is covered by each ploidy state.

6.  Add a flag when any ploidy state, other that "Normal diploid", covers over a threshold percentage of the total chromosome arm.

```{r}
#| label: chr-arm-percent-threshold
#| include: FALSE

chr_arm_percent_threshold <- 75

```

The threshold for flagging a chromosome arm level CNV event was set at
`r chr_arm_percent_threshold`%.
This was decided due to the 

## Orthogonal testing

### Karyotype analysis

Karyotype analysis was performed at the North West GLH (Liverpool site).

### Fluorescence In-Situ Hybridisation

Fluorescence In-Situ Hybridisation (FISH) analysis was performed at the
North West GLH (Liverpool site) using the Zytolight 1p36/1q25 dual
colour probes (ZytoVision).

### SNP array

Haemato-oncology samples were tested using the Illumina Global Screening
Array-24+ 3.0 assay at the North West GLH (Manchester site).

### Methylation array

Samples were processed on the Illumina Infinium Methylation EPIC Bead
Chip platform at the North West GLH and data were analysed using the
University of Heidelberg Central Nervous System methylation classifier
[@capper2018].

### Whole genome sequencing

Whole genome sequencing (WGS) was performed by Illumina Laboratory
Services in Cambridge. Further details are provided in
[DOC6567]{custom-style="Strong"}.

## Validation analysis

Analysis of the validation data was performed using R (4.4.1).

# Results

## Distribution of ploidy state regions

```{r}
#| label: cohort-data
#| include: FALSE

pansolid_unannotated_files <- list.files(paste0(config::get("data_folderpath"),
                 "validation/DOC6791_chromosome_arms/raw/pansolid_ngs/unannotated/"),
                 "Results_SNVs_Indels-.*.xlsx",
                 full.names = TRUE)

pansolid_ploidy_data <- pansolid_unannotated_files |> 
  map(\(pansolid_unannotated_files) 
      read_ploidy_sheet(filepath = pansolid_unannotated_files,
                        sheetname = get_sheetname(filepath = pansolid_unannotated_files,
                                                  sheet_regex = "Ploidy\\sRegion.*"))) |> 
  list_rbind()

gain_names <- c("(1,3)", "(0,4)", "(0,3)", "(2,4)", 
                "Whole genome duplication", "Duplication", 
                "(2,3)")

del_names <- c("Bi-allelic deletion", "Deletion")

pansolid_ploidy_data_long <- pansolid_ploidy_data |> 
  extract_pansolid_cnv_coordinates(cnv_coord_col = region) |> 
    format_chromosome_decimals() |> 
    factorise_chromosome() |> 
    add_chromosome_arms() |> 
    calculate_region_size() |> 
  mutate(name_group = case_when(
    name == "Normal diploid" ~"Normal diploid",
    name == "Uniparental disomy" ~"Uniparental disomy",
    name %in% gain_names ~"Gain",
    name %in% del_names ~"Deletion"
  ),
  name_group = factor(name_group, levels = c("Normal diploid",
                                             "Deletion",
                                             "Gain",
                                             "Uniparental disomy")))
```

The ploidy state results for 
`r length(unique(pansolid_ploidy_data$labno))`
samples were analysed by the chromosome arm quantification method. 
The overall distribution of results is shown in @fig-ploidy-state-histogram, which shows a clear sub-group of ploidy state calls where a single ploidy state call accounts for over 95% of the chromosome arm.
This indicates that the threshold of 
`r chr_arm_percent_threshold`% is an appropriate threshold for discriminating large chromosome events.

```{r}
#| label: fig-ploidy-state-histogram
#| echo: FALSE

pansolid_data_for_plot <- pansolid_ploidy_data_long |> 
  filter(name_group != "Normal diploid") |> 
  mutate(percent_chr_arm_round = round(percent_chr_arm, 1)) |> 
  filter(percent_chr_arm_round >= 1)

fig_ploidy_state_histogram <-  ggplot(pansolid_data_for_plot, 
                                      aes(x = percent_chr_arm_round, 
                                          y = )) +
  geom_histogram(binwidth = 2, aes(fill = name_group), colour = "black") +
  scale_fill_manual(values = c("blue", "green", "orange")) +
  geom_vline(xintercept = chr_arm_percent_threshold, linetype = "dashed") +
  scale_x_continuous(limits = c(0, 100),
                     breaks = seq(0, 100, by = 10)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "Percentage of chromosome arm covered by call",
       y = "Number of ploidy state calls",
       fill = "",
       title = paste0("Size distribution for ",
                      nrow(pansolid_data_for_plot),
                      " ploidy state calls in ",
                      length(unique(pansolid_ploidy_data_long$labno)),
                      " samples"),
       subtitle = "Normal diploid ploidy states removed. Bin size: 2%")

fig_ploidy_state_histogram

```

The `r chr_arm_percent_threshold`% threshold was used to flag when a chromosome arm was impacted by a large CNV event. 
An example is shown in @tbl-example-full.

```{r}
#| label: analyse-ploidy-data
#| include: FALSE

pansolid_ploidy_results <- pansolid_ploidy_data |>
  extract_pansolid_cnv_coordinates(cnv_coord_col = region) |> 
  format_chromosome_decimals() |> 
  factorise_chromosome() |> 
  add_chromosome_arms() |> 
  calculate_region_size() |> 
  add_chr_result_flag(percent_threshold = chr_arm_percent_threshold) |> 
  format_arm_ploidy_tbl()
  
```

{{< pagebreak >}}

```{r}
#| label: tbl-example-full
#| tbl-cap: "Example of quantitative method output for sample 24017214"
#| echo: FALSE

example_tbl_full <- pansolid_ploidy_results |> 
  filter(labno_suffix_worksheet == "24017214_WS145856") |>  
  ungroup() |> 
  select(chromosome_arm, result_flag, full_string)

knitr::kable(example_tbl_full)

```

{{< pagebreak >}}

```{r}
#| label: tbl-example-summary
#| tbl-cap: "Example of summarised output for sample 24017214"
#| echo: FALSE

knitr::kable(example_tbl_full |> 
               filter(result_flag != ""))

```

## Comparison to 1p19q FISH

```{r}
#| label: fish-comparison
#| echo: FALSE

fish_results <- read_excel(path = paste0(config::get("data_folderpath"),
                 "validation/DOC6791_chromosome_arms/raw/fish/",
                 "fish_result_summary.xlsx"),
                 col_types = c("text", "text", "text", "text")) |> 
  pivot_longer(cols = c("1p", "19q"),
               names_to = "chromosome_arm",
               values_to = "fish_result")

val_1p19q <- pansolid_ploidy_results |> 
  filter(chromosome_arm %in% c("1p", "19q")) |> 
  mutate(labno = str_extract(string = labno_suffix_worksheet,
                             pattern = "(\\d{8})_WS\\d{6}",
                             group =1)) |> 
  select(labno_suffix_worksheet, labno,
         chromosome_arm, result_flag, full_string) |> 
  arrange(labno_suffix_worksheet, desc(chromosome_arm)) |> 
  inner_join(fish_results,
            join_by("labno" == "pansolid_labno",
                    "chromosome_arm" == "chromosome_arm")) |> 
  select(name, labno, chromosome_arm, result_flag, fish_result,
         full_string)

knitr::kable(val_1p19q)

```

8 results are in agreement between the two methods.

For 3 results, FISH reports a 19q deletion whilst PanSolid shows terminal 19q deletions with less than `r chr_arm_percent_threshold`% of 19q deleted.
This presumably is a false positive result on FISH.

One result does not agree because 1p and 19q are a mixture of LOH and deletion calls on PanSolid.

## Comparison to whole genome sequencing

### 1q19p gain: sample 23001331

Sample 23001331 was tested. 
A 1p19q codeletion was detected on the previous CNV detection method (DCO4914) (report reference: 23001331.123).

No 1p19q codeletion was detected on WGS (R25-0YP8-1), however a 1p
which had a 1q19p gain on WGS, in order to check whether this would be incorrectly interpreted as a 1p19q loss.
Testing of this sample on the previous CNV detection method (see DOCxxxx) had 

```{r}
#| label: tbl-sample-23001331
#| echo: FALSE

tbl_sample_23001331 <- pansolid_ploidy_results |> 
  filter(labno_suffix_worksheet == "23001331_WS156025") |> 
  ungroup() |> 
  filter(chromosome_arm %in% c("1p", "1q", "19p", "19q")) |> 
  select(chromosome_arm, result_flag, full_string)

knitr::kable(tbl_sample_23001331)

```

### Gain of 7 and loss of 10

```{r}
#| label: fig-plus-7-minus-10
#| include: FALSE

plus7_loss10_samples <- c("25024755_WS152973", 
                          "25029510_WS153752",
                          "25032698_WS154250")

x <- pansolid_ploidy_results |> 
  filter(labno_suffix_worksheet %in% plus7_loss10_samples &
           chromosome_arm %in% c("7p", "7q",
                                 "10p", "10q")) 

```

## Comparison to G-band karyotype

```{r}
#| label: karyotype-samples
#| include: FALSE

karyotype_samples <- read_excel(path = paste0(config::get("data_folderpath"),
                                              "validation/DOC6791_chromosome_arms/",
                                              "karyotype_samples_to_test.xlsx")) |> 
  janitor::clean_names()

```

5 samples with known hyperdiploid or hypodiploid karyotypes from G-banding analysis were tested on PanSolid NGS.
The karyotype results are shown in @tbl-karyotype-samples.

```{r}
#| label: tbl-karyotype-samples
#| tbl-cap: "Samples with karyotype results"
#| echo: FALSE

karyotype_tbl <- karyotype_samples |> 
  select(sample_id, folder_name, karyotype) |> 
  rename(`Lab number` = sample_id,
         `Karyotype number` = folder_name,
         `Karyotype` = karyotype)

knitr::kable(karyotype_tbl)

```

```{r}





pansolid_ploidy_results |> 
  filter(grepl("24062890", labno_suffix_worksheet) &
           result_flag != "") |>  view()


```



## Repeat testing

3 DNA samples from Schwannoma tumours (24018518, 24045597 and 24054282)
were tested on 3 different worksheets to assess the impact of repeat
variation. These were the same DNA samples used for the validation of
detection of deletions ([DOC6567]{custom-style="Strong"}).

### Uncertainty of measurement

## Quality control

## Limit of detection

# Discussion

This new method is an improvement on the previous NGS detection method for 1p19q, as it does not rely on normalising dosage to a single chromosome arm.
This is demonstrated by 23001331.

## Limitations

Bone marrow samples were tested because only bone marrow samples get karyotyping. Karyotyped samples from FFPE blocks were not available.

The quantification of chromosome arm percent will not sum to 100%, because the PanSolid enrichment does not cover the entire breadth of every chromosome arm.

Summing ploidy state regions across a chromosome arm can merge regions with separate biological causes. If a chromosome arm has multiple independent deleted regions, these will be summed to give the impression that a single deletion is impacting the chromosome arm.

## Extrapolation to other copy number variants

This validation has focussed on a discrete list of chromosome-level
CNVs. The sensitivity and specificity of the PanSolid pipeline for
detecting the chromosome level CNVs within the validation data can be
extrapolated to all chromosome level CNVs, provided that the CNV has a
size greater than X Megabases.

# Review of Test Requirements

# Requirements for Clinical Implementation

# Conclusion

# References

::: {#refs}
:::
