---
title: "DOC6791: Validation of Detecting Chromosome Level Copy Number Variants with the PanSolid CLC pipeline"
format: 
  docx:
    reference-doc: north_west_glh_document_template.docx
    link-citations: true
toc: true
toc-title: "Table of Contents"
bibliography: pansolid_cnv_references.bib
csl: harvard-manchester-metropolitan-university.csl
fig-align: "left"
fig-wdith: 6
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: packages-and-functions
#| include: FALSE

library(tidyverse)
library(janitor)

```

# Authors

Author: Joe Shaw (CS20980)

Project Lead(s): Joe Shaw, Elizabeth Sollars, Eleanor Baker, Helene
Schlecht, George Burghel

Date: `r Sys.Date()`

Author contact: joseph.shaw2\@mft.nhs.uk; joseph.shaw3\@nhs.net

# Acronyms

**ATRX**: α-thalassemia/mental retardation syndrome X-linked

**cIMPACT-NOW:** consortium to Inform Molecular and Practical Approaches
to CNS Tumour taxonomy-Not Official WHO

**CLC**: there is no official acronym for CLC, although unofficially it was referred to as "Cake Loving Company" by CLC employees [@sollars2025]

**DNA**: DeoxyriboNucleic Acid

**FFPE**: Formalin-Fixed Paraffin Embedded

**FISH**: Fluorescent In-Situ Hybridisation

**GLH**: Genomic Laboratory Hub

**IDH**: isocitrate dehydrogenase

**IHC**: Immuno-HistoChemistry

**LOH**: Loss Of Heterozygosity

**MGMT**: O6-Methylguanine-DNA methyltransferase

**NCC**: Neoplastic Cell Content

**NGS**: Next Generation Sequencing

**NHS**: National Health Service

**PCR:** Polymerase Chain Reaction

**SNP:** Single Nucleotide Polymorphism

**TERT**: Telomerase Reverse Transcriptase

**WGS**: Whole Genome Sequencing

**WHO**: World Health Organisation

# Introduction

## Purpose

The purpose of this document is to describe the validation of detection
of chromosome level copy number variants (CNVs) in solid tumour samples
with the PanSolid v2 QIAseq enrichment panel using the PanSolid CLC
pipeline (v2.4) at the North West Genomic Laboratory Hub (GLH).
Detection of specific CNVs in different cancer types is required by the
most recent update of the
[NHS Genomic Test Directory for
Cancer](https://www.england.nhs.uk/publication/national-genomic-test-directories/).

## Target copy number variants

### Neurological tumours

Co-deletion of 1p and 19q, as determined by PCR-based LOH testing, was
shown to be associated with a good response to chemotherapy in
oligodendrogliomas [@cairncross1998]. The biological mechanism for this
observation remains unknown.

Later cytogenetic studies using Giemsa-banding and FISH showed that the
co-deletion of 1p and 19q was due to a recurrent translocation between
the two chromosomes. A derivative der(1;19)(q10;p10) chromosome was
observed within glioma cells, whilst the reciprocal der(1;19)(p10;q10)
chromosome is absent and presumably lost [@jenkins2006; @griffin2006].

In the 2021 edition of the World Health Organisation Guidelines,
detection of 1p19q codeletion is required for classification of a tumour
as "Oligodendroglioma, IDH-mutant, and 1p/19q-codeleted" [@louis2021].
Critically, 1p19q codeletion is a key biomarker for distinguishing
between oligodendrogliomas, which contain the codeletion, and
astrocytomas, which do not [@masui2016].

Multiple techniques can be used to detect 1p19q codeletions and there is
no gold-standard method [@brandner2022].

In the most recent update of the cIMPACT-NOW guidelines for meningiomas,
whole chromosome copy number analysis is recommended for tumour grading,
with at least two losses of chromosomes 6, 10, 14 or 18 sufficient to
categorise a tumour as molecular high-risk [@sahm2025].

### Malignant melanoma

The purpose is to distinguish benign nevi from malignant melanomas.

References: @ferrara2016; @gerami2009

### Renal cell carcinoma

Specific chromosome gains and losses are associated with subtypes of
renal cancer.

Renal cell carcinoma: loss of 3p (including VHL)

Papillary renal cell carcinoma: gains of 7 and 17

References: @liu2020, @klatte2009, @chen2009

## PanSolid enrichment

```{r}
#| label: enrichment-targets
#| echo: FALSE

target_df <- read_csv(paste0(config::get("data_folderpath"),
                             "validation/DOC6283_amplifications/bed_files/",
                             "PanSolidv2_GRCh38_noalt_BED.csv")) |> 
  clean_names() |> 
  mutate(chromosome = factor(chromosome,
                                levels  = c(seq(1:22), "X")))

targets_per_chromosome <- target_df |> 
  group_by(chromosome) |> 
  count() 

ggplot(targets_per_chromosome, aes(x = chromosome, y = n)) +
  geom_col() +
  theme_bw() +
  labs(x = "Chromosome", y = "Number of targets",
       title = "Distribution of targets on PanSolid enrichment")

```

## Requirements and Performance Specification

The requirements for this new method are as follows:

-   The test must have a clear visual presentation of results
-   The test must provide a quantitative result for whether a chromosome
    or chromosome arm is impacted by a gain or loss.
-   The combined analysis of the PanSolid CLC pipeline with manual
    review by a clinical scientist should sensitivity and specificity
    values above 90% when compared with orthogonal testing.
-   The test must have clear quality metric thresholds for deciding
    sample quality

# Methods

## Samples



## Next generation sequencing

Next generation sequencing on the PanSolid v2 enrichment was performed
for all samples.

## PanSolid CLC bioinformatics pipeline

The PanSolid CNV pipeline was built in CLC Genomics Workbench and has been described in previous validation documents.

### Visualisation and quantification

The outputs of the PanSolid CNV pipeline are visualised at the whole-genome level via interactive plots, which were generated using the plotly package by Elizabeth Sollars and implemented during the validation of deletion and loss of heterozygosity detection (DOCxxxx).
In addition, scientists in the somatic cancer team requested an quantitative way of alerting them when a CNV impacting a whole chromosome arm or arms was present, such as the 1p19q co-deletion.
The following method 

1.  Load the ploidy regions from the "Ploidy Regions" tab of the "Results_SNVs_Indels" Excel file in the "Unannotated_Files" folder of each worksheet.

2.  Annotate each region by the chromosome arm that the region is on.

3.  Calculate the size of the region by subtracting the start coordinate from the end coordinate.

4.  Sum the total length, in base pairs, of ploidy regions for a particular ploidy state, such as "Deletion", on each chromosome arm.

5.  Calculate the percentage of the entire chromosome arm that is covered by each ploidy state.

6.  Add a flag when any ploidy state, other that "Normal diploid", covers over 90% of the total chromosome arm.

## Orthogonal testing

### Karyotype analysis

Karyotype analysis was performed at the North West GLH (Liverpool site).

### Fluorescence In-Situ Hybridisation

Fluorescence In-Situ Hybridisation (FISH) analysis was performed at the
North West GLH (Liverpool site) using the Zytolight 1p36/1q25 dual
colour probes (ZytoVision).

### SNP array

Haemato-oncology samples were tested using the Illumina Global Screening
Array-24+ 3.0 assay at the North West GLH (Manchester site).

### Methylation array

Samples were processed on the Illumina Infinium Methylation EPIC Bead
Chip platform at the North West GLH and data were analysed using the
University of Heidelberg Central Nervous System methylation classifier
[@capper2018].

### Whole genome sequencing

Whole genome sequencing (WGS) was performed by Illumina Laboratory
Services in Cambridge. Further details are provided in
[DOC6567]{custom-style="Strong"}.

## Validation analysis

Analysis of the validation data was performed using R (4.4.1).

# Results

## Neurological tumours

### 1p19q codeletion samples

X samples were tested, which had 1p19q codeletion results from FISH or WGS, or both.

### 1q19p gain

Sample 23001331 was tested which had a 1q19p gain, in order to check whether this would be incorrectly interpreted as a 1p19q loss.

A sample from a tumour with chromothripsis was tested.

## Malignant melanoma

## Renal cell carcinoma

## Comparison to G-band karyotype

5 samples with known hyperdiploid or hypodiploid karyotypes from G-banding analysis were tested on PanSolid NGS.

## Repeat testing

3 DNA samples from Schwannoma tumours (24018518, 24045597 and 24054282)
were tested on 3 different worksheets to assess the impact of repeat
variation. These were the same DNA samples used for the validation of
detection of deletions ([DOC6567]{custom-style="Strong"}).

### Uncertainty of measurement

## Limit of detection

# Discussion

## Limitations

Bone marrow samples were tested because only bone marrow samples get karyotyping. Karyotyped samples from FFPE blocks were not available.

The quantification of chromosome arm percent will not sum to 100%, because the PanSolid enrichment does not cover the entire breadth of every chromosome arm.

## Extrapolation to other copy number variants

This validation has focussed on a discrete list of chromosome-level
CNVs. The sensitivity and specificity of the PanSolid pipeline for
detecting the chromosome level CNVs within the validation data can be
extrapolated to all chromosome level CNVs, provided that the CNV has a
size greater than X Megabases.

# Review of Test Requirements

# Requirements for Clinical Implementation

# Conclusion

# References

::: {#refs}
:::
