---
title: "PanSolidv2 CNV Audit"
author: "Joe Shaw, Clinical Scientist (CS20980)"
date: today
date-format: "DD/MM/YYYY"
format: html
editor: visual
---

```{r}
#| label: packages
#| include: FALSE

library(tidyverse)
library(readxl)
library(here)

```

Copy number variant results are collated using the collate_pansolid_cnv_service_data.R script.

Results are updated using the update_pansolid_cnv_service_data.R script.

```{r}
#| label: functions
#| include: FALSE

source(here::here("functions/cnv_functions.R"))

```

```{r}
#| label: load-collated-data
#| include: FALSE

amp_gene_results <- read_csv(here::here("data/live_service_collated_data/live_service_amp_gene_results_collated.csv"))

std_dev_results <- read_csv(here::here("data/live_service_collated_data/live_service_std_dev_results_collated.csv"))

percent_138_results <- read_csv(here::here("data/live_service_collated_data/live_service_percent_138_results_collated.csv"))

pos_cnv_results <- read_csv(here::here("data/live_service_collated_data/live_service_pos_cnv_results_collated.csv"))

```

```{r}
#| label: join-collated-data
#| include: FALSE

fold_change_threshold <- 2.8

pansolid_cnv_collated_data <- amp_gene_results |> 
  left_join(std_dev_results |> 
              select(filepath, st_dev_signal_adjusted_log2_ratios),
            by = "filepath") |> 
  mutate(result = case_when(
    
    max_region_fold_change >= fold_change_threshold ~"Amplification detected",
    
    max_region_fold_change < fold_change_threshold ~"No amplification detected")) |> 
  left_join(percent_138_results |> 
              select(filepath, percent_whole_panel_covered_at_138x),
            by = "filepath")

```

```{r}
#| label: count-colorectal-referrals
#| include: FALSE

pansolidv2_worksheets <- read_excel(here::here("data/pansolid_live_service_worksheets.xlsx"))

worksheet_list <- list(pansolidv2_worksheets$worksheet)

s_drive_filepaths <- worksheet_list |> 
  map(\(worksheet_list) get_annotated_filepaths(worksheet_list)) |> 
  flatten()

s_drive_file_df <- tibble(
  filepath = unlist(s_drive_filepaths)) |> 
  mutate(filename = str_extract(string = filepath, 
                         pattern = str_replace(string = pansolidv2_excel_regex, 
                                               pattern = "\\^", 
                                               replacement = "")),
         labno = str_extract(string = filename, 
                             pattern = "WS\\d{6}_(\\d{6,8})_",
                             group = 1))

colorectal_filepaths <- grep(pattern = "CRC", x = s_drive_file_df$filepath, 
                             ignore.case = TRUE,
                             value = TRUE)

colorectal_df <- s_drive_file_df |> 
  filter(filepath %in% colorectal_filepaths)
  
erbb2_crc_results <- amp_gene_results |> 
  mutate(filename = str_extract(string = filepath, 
                         pattern = str_replace(string = pansolidv2_excel_regex, 
                                               pattern = "\\^", 
                                               replacement = ""))) |> 
  filter(gene == "ERBB2" & max_region_fold_change >= fold_change_threshold) |> 
  filter(filename %in% colorectal_df$filename)

live_weeks <- round(as.numeric(difftime(today(), date("2024-04-08"), units = "weeks")), 1)

```

# Results

Total samples: `r length(unique(pansolid_cnv_collated_data$labno))`

Colorectal cancer samples: `r length(colorectal_filepaths)`

*ERBB2* amplifications in colorectal cancer samples: `r nrow(erbb2_crc_results)`

Weeks PanSolidv2 has been live: `r live_weeks`

Here's the summary of amplifications detected for samples passing the quality filter (signal-adjusted noise metric below 1).

```{r}
#| label: summary-gene-amps-filter
#| echo: FALSE

summary_table_filter <- pansolid_cnv_collated_data |> 
  filter(st_dev_signal_adjusted_log2_ratios < 1) |> 
  group_by(gene, result) |> 
  count() |> 
  pivot_wider(names_from = result, values_from = n, values_fill = 0) |> 
  select(gene, `Amplification detected`, `No amplification detected`) |> 
  arrange(desc(`Amplification detected`))

knitr::kable(summary_table_filter)

```

{{< pagebreak >}}

And here's the overall view of gene fold changes.

```{r}
#| label: amps-plots
#| echo: FALSE

gene_fold_changes_plot <- ggplot(pansolid_cnv_collated_data |> 
                                   filter(st_dev_signal_adjusted_log2_ratios < 1), 
                                 aes(x = gene, y = max_region_fold_change)) +
  geom_jitter(pch = 21, width = 0.4, alpha = 0.5, aes(fill = result)) +
  scale_fill_manual(values = c(safe_red, safe_blue)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "", y = "Maximum region fold change",
       caption = "Data passing signal-adjusted noise QC") +
  geom_hline(yintercept = fold_change_threshold, linetype = "dashed") 

gene_fold_changes_plot

```

## Quality Metrics

We can also look distribution of the "signal-adjusted noise" (standard deviation of signal-adjusted log2 ratios) and percentage of the panel covered to at least 138X.

```{r}
#| label: qc-plot
#| echo: FALSE

noise_plot <- ggplot(pansolid_cnv_collated_data |> 
         filter(!duplicated(labno)), 
       aes(x = st_dev_signal_adjusted_log2_ratios, y = )) +
  geom_histogram(binwidth = 0.1) +
  theme_bw() +
  scale_x_continuous(breaks = seq(0, 3.5, by = 0.5)) +
  labs(y = "Number of samples")

noise_plot

percent_138_plot <- ggplot(pansolid_cnv_collated_data |> 
         filter(!duplicated(labno)), 
       aes(x = percent_whole_panel_covered_at_138x, y = )) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  labs(y = "Number of samples")

percent_138_plot

```

```{r}
#| label: qc-plot-by-worksheet
#| echo: FALSE

ggplot(pansolid_cnv_collated_data |> 
         filter(!duplicated(labno)), 
       aes(x = worksheet, y = st_dev_signal_adjusted_log2_ratios)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 4)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "", y = "Signal-adjusted noise")

```

```{r}
#| label: 138x-plot-by-worksheet
#| echo: FALSE

ggplot(pansolid_cnv_collated_data |> 
         filter(!duplicated(labno)), 
       aes(x = worksheet, y = percent_whole_panel_covered_at_138x)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "", y = "Percent panel covered to 138X")

```

{{< pagebreak >}}

## Amplification Details

These are the samples with amplifications detected.

```{r}
#| label: view-amps
#| echo: FALSE

amps_detected <- pansolid_cnv_collated_data |> 
  filter(st_dev_signal_adjusted_log2_ratios < 1 &
           result == "Amplification detected") |> 
  select(worksheet, labno, gene, max_region_fold_change, 
         st_dev_signal_adjusted_log2_ratios) |> 
  arrange(worksheet, gene)

knitr::kable(amps_detected)

```
