---
title: "Fold Change Threshold"
author: "Joe Shaw"
date: today
date-format: "DD/MM/YYYY"
format: html
editor: visual
---

```{r}
#| label: libraries
#| include: FALSE

library(tidyverse)

```

## Fold change

The CLC fold change is the observed coverage divided by the expected coverage.

This can be simplified to the number of target copies per cell divided by the expected copies per cell (2).

If there are 3 target copies per cell, the fold change is 1.5 (3/2).

## Tumour cell content

The variable tumour cell content of samples makes things trickier because the sample is not pure.

To calculate how this impacts the fold change, it is useful to think about the number of target copies per 100 cells.

If 50% of the sample has tumour cells with 10 target copies per cell, you have to work out the contribution of the tumour cells and normal cells to the total number of target copies in the sample.

Here is the calculation as a function:

```{r}
#| label: calc-fold-change
#| include: TRUE

# The number of copies of a reference locus per cell (2) multiplied by 100 cells.
ref_sample_copies <- 200

calculate_fold_change <- function(tcc_percent, sample_target_copies_per_tumour_cell) {
  
  tcc_fraction <- tcc_percent / 100
  
  sample_target_copies_in_tumour_cells <- (100 * tcc_fraction) * sample_target_copies_per_tumour_cell
  
  sample_target_copies_in_normal_cells <- (100 * (1-tcc_fraction)) * 2
  
  sample_total_target_copies <- sample_target_copies_in_tumour_cells + sample_target_copies_in_normal_cells
  
  fold_change <- sample_total_target_copies / ref_sample_copies
  
  return(fold_change)
  
}

calculate_fold_change(50, 10)

```

## Fold change threshold

The CLC pipeline uses a fold change threshold of 3 for calling copy number variants.

We can rearrange the previous function to allow us to calculate the number of target copies required to give a fold change of 3 at each tumour cell content value. 

```{r}
#| label: calc-target-copies
#| include: TRUE

calculate_target_copies <- function(fold_change, tcc_percent) {
  
  tcc_fraction <- tcc_percent / 100
  
  sample_total_target_copies <- fold_change * ref_sample_copies

  sample_target_copies_in_tumour_cells <- sample_total_target_copies - ((100 * (1-tcc_fraction)) * 2)

  sample_target_copies_per_tumour_cell <- sample_target_copies_in_tumour_cells / (100 * tcc_fraction)
  
  return(sample_target_copies_per_tumour_cell)
  
}

```

This then shows us which combinations of tumour cell content and target copy number lie above the threshold, and which lie beneath it.

```{r}
#| label: draw-plot
#| echo: FALSE

df <- data.frame(tcc = seq(0, 100, by = 0.1)) |> 
  mutate(target_copies = calculate_target_copies(fold_change = 3,
                                                 tcc_percent = tcc))
fold_change_plot <- df |> 
  filter(target_copies <= 100) |> 
  ggplot(aes(x = target_copies, y = tcc)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(x = "Target copies per cell", y = "Tumour cell content (%)",
       title = "Detection limit of CLC pipeline",
       subtitle = "Line shows detection limit using a fold change threshold of 3") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0,100,10),
                     minor_breaks = FALSE) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0,100,10),
                     minor_breaks = FALSE) +
  geom_text(aes(x = 30, y = 30), size = 4, label = "Detected") +
  geom_text(aes(x = 10, y = 10), size = 4, label = "Not detected")

fold_change_plot

```
